Pleiotropy of expression quantitative trait loci increases with the number of regulated genes, active cell types and tissues and bound transcription factors

### 2024-11-10 ms/ms01_abstract.tex
First commit after acceptation

%\leadauthor{Scientist}

\title{Pleiotropy of expression quantitative trait loci increases with the number of regulated genes, active cell types and tissues and bound transcription factors}
%\shorttitle{Pleiotropy and gene regulatory variants}

\author[1,*]{Aitor González\,\orcidlink{0000-0002-8402-1655}}

%\author[1,\Letter]{Aitor González}
\affil[1]{Aix Marseille Univ, INSERM, TAGC, 13288 Marseille, France}
\affil[*]{Corresponding author: aitor.gonzalez@univ-amu.fr}
% \orcidlink{0000-0002-8402-1655}
%\affil[1]{Aix Marseille Univ, INSERM, TAGC, 13288 Marseille, France}
\date{}

\maketitle

\begin{abstract}

Pleiotropic genetic variants affecting several traits are relatively common in the genome as shown by the increasing number of genome-wide association studies (GWAS).
%
Frequent variants are often located in non-coding regions and are likely gene regulatory variants.
%
In this work, I investigate the link between frequent gene regulatory variants and pleiotropy.

% Verify numbers
First I have computed the colocalization probability of 418 GWAS and 127 eQTL studies and I have observed 10,334 GWAS/eQTL pairs with at least one colocalized variants with a probability above 0.8.
%SELECT count(DISTINCT concat(`gwas_id`, `eqtl_id`)) FROM `gwas2eqtl_modelgwas2eqtl` where `PP_H4_abf`>=0.8; 10334
%Pourcentage of GWAS
%
I have categorized the GWAS traits into 96 traits and categories to define pleiotropic as belonging at least to two of these categories.
%
Pleiotropic variants are associated to more eQTL genes (eGenes) and more eQTL tissues (etissues) but egenes do also correlate with more traits.
%
These observations could be partially explained, because pleitropic variants are enriched in splicing and missense variants and bind more transcription factors.
%
These variants with the GWAS traits, egenes and etissues can be explored in a public website.

In conclusion, our work suggest that pleiotropy of gene regulatory variants arise from an increase of bound transcription factors, target egenes and etissues and severity of molecular effect consequences.

\end{abstract}

% Keywords
% eGenes; eQTLs

>>>>

\begin{abstract} % abstract

    \parttitle{Background} %if any
    %
    Pleiotropic genetic variants affecting several traits are relatively common in the genome as shown by the increasing number of genome-wide association studies (GWAS).
%
    Complex disease variants are often located in non-coding regions and are likely gene regulatory variants.
%
    In this work, I investigate the link between frequent expression quantitative trait loci and pleiotropy.

    \parttitle{Results} %if any
% Verify numbers
    First I have computed the colocalization probability of 417 GWAS and 127 eQTL studies and I have observed 10,334 GWAS/eQTL pairs with at least one colocalized variants with a probability above 0.8.
%SELECT count(DISTINCT concat(`gwas_id`, `eqtl_id`)) FROM `gwas2eqtl_modelgwas2eqtl` where `PP_H4_abf`>=0.8; 10334
%Pourcentage of GWAS
%
    I have categorized the GWAS traits into 96 traits and categories to define pleiotropic as belonging at least to two of these categories.
%
    Pleiotropic variants are associated to more eQTL genes (eGenes) and more eQTL tissues (etissues) but egenes do also correlate with more traits.
%
    These observations could be partially explained, because pleitropic variants are enriched in splicing and missense variants and bind more transcription factors.
%
    These variants with the GWAS traits, egenes and etissues can be explored in a public website.

    \parttitle{Conclusions} %if any
    In conclusion, our work suggest that pleiotropy of gene regulatory variants arise from an increase of bound transcription factors, target egenes and etissues and severity of molecular effect consequences.
    %
\end{abstract}

### 2024-11-10 ms/ms05_methods.tex
First commit after acceptation

I used these data sets here: eQTLs from the eQTL Catalogue \citep{2021.Alasoo.Kerimov}, GWAS variants from the IEU OpenGWAS project \citep{2021.Marcora.Lyon}, chromatin immunoprecipitation (ChIP) transcription factors peaks from the ReMap database \citep{2021.Ballester.Hammal}, and UCSC annotation data.
%
I explore the data using the UCSC browser (\url{http://genome.ucsc.edu}) \citep{2021.Kent.Lee}, and OMIM database (\url{https://omim.org/}).
%
Colocalization and analysis pipelines were implemented with Snakemake (Supplementary Figure S1) .

@@

Full association data were downloaded from OpenGWAS and converted to hg38 coordinates using Picard and Crossmap \citep{2021.Marcora.Lyon,Picard2019toolkit,2013.Wang.Zhao}.
%
Top hit variants from the corresponding GWAS were download based on p-value threshold 0.8, clumping parameter r$^2$=0.1 and a 1 Mb clumping window.
%
eQTL and GWAS variants in a total window with diameter 1 Mb around the top hits variants.
%
Missing minor allele frequency variants (MAF) were retrieved from the European population in the 1000 genome database \citep{2015.Abecasis.Auton}.
%
Variants were kept if MAF was strictly between 0 and 1, the variants were not duplicated and there was no missing data.
%
Colocalization between eQTL and GWAS variants was tested using the "coloc.abf" function of CRAN coloc for each window, each eQTL sample and each GWAS.
%
The colocalization and analysis pipelines can be found here: \url{https://github.com/aitgon/gwas2eqtl} and \url{https://github.com/aitgon/gwas2eqtl_pleiotropy}.

>>>>

I used these data sets here: eQTLs from the eQTL Catalogue \cite{2021.Alasoo.Kerimov}, GWAS variants from the IEU OpenGWAS project \cite{2021.Marcora.Lyon}, chromatin immunoprecipitation (ChIP) transcription factors peaks from the ReMap database \cite{2021.Ballester.Hammal}, and UCSC annotation data.
%
I explore the data using the UCSC browser (\url{http://genome.ucsc.edu}) \cite{2021.Kent.Lee}, and OMIM database (\url{https://omim.org/}).
%
Colocalization and analysis pipelines were implemented with Snakemake (Supplementary Figure S1) .

@@

Full association data were downloaded from OpenGWAS and converted to hg38 coordinates using Picard and Crossmap \cite{2021.Marcora.Lyon,Picard2019toolkit,2013.Wang.Zhao}.
%
Top hit variants from the corresponding GWAS were download based on p-value threshold 0.8, clumping parameter r$^2$=0.1 and a 1 Mb clumping window.
%
eQTL and GWAS variants in a total window with diameter 1 Mb around the top hits variants.
%
Missing minor allele frequency variants (MAF) were retrieved from the European population in the 1000 genome database \cite{2015.Abecasis.Auton}.
%
Variants were kept if MAF was strictly between 0 and 1, the variants were not duplicated and there was no missing data.
%
Colocalization between eQTL and GWAS variants was tested using the "coloc.abf" function of CRAN coloc for each window, each eQTL sample and each GWAS.
%
The colocalization and analysis pipelines can be found here: \url{https://github.com/aitgon/gwas2eqtl} and \url{https://github.com/aitgon/gwas2eqtl_pleiotropy}.
